Ricci curvature of the rooted phylogenetic subtree-prune-regraft graph

### 2015-02-10 curvature.tex
curvature lower bound

>>>>

Note that this bound is tight and has been verified computationally for $n \le 7$.

Next, we bound the minimum curvature of two adjacent trees.

\begin{lemma}
	The curvature of the uniform random walk between two adjacent trees with $n$ leaves is at least
	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
\end{lemma}
\begin{proof}
	Let $T$ and $S$ be two adjacent trees.
	We can bound the mass transport cost from above by $1 + 2m$, where $m$ is the maximum proportion of mass that can not be moved between adjacent trees. 
	This gives a lower bound of $1 - (1 + 2m) / 1) = -2m$ on the curvature.
	The mass transport cost is maximized (and therefore curvature minimized) by a combination of two effects: trees that must be paired at a distance $>1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.  
	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
	By Corollary~\ref{cor:paired_neighbors}, there are at most $o := \degree{T} -2ka - 2(i-1)$ paired trees.
	Then, let $S$ be the ladder tree with degree $3n^2 - 13n + 14$ and $T$ be the adjacent tree constructed by moving the lower $\floor{\frac{n}{2}}$ leaves of $S$ to the root.
	$T$ has degree at most $3.5n^2 -15n + 16$.
	There are thus $2ka + 2(i-1) = 2\left(\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n$ unpaired neighbors, the maximum possible.
	Then the maximum $m$ is:
	$$\frac{\frac{1}{2}n^2 - n}{3.5n^2 - 15n + 16}.$$
	The claim follows from multiplying this value by $-2$.
\end{proof}

We further observe that the limit of our curvature lower bound is $-\frac{2}{7}$.
Exhaustive tests with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
Moreover, the SPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees\todo{cite one of the Ollivier papers}.
This directly leads to the following corollary:

\begin{corollary}
	The curvature of the uniform random walk between two phylogenetic trees is at least $-\frac{2}{5}$.
\end{corollary}

Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
